# Introduction {#intro}

The entities R operates on are technically known as **objects**. There are two types of objects:

1. Data structures

2. Functions

# Data structures in base R

There are five data types in R

  1. Atomic vector
    
  2. Matrix
    
  3. Array
    
  4. List
    
  5. Data frame
    
## Atomic vectors

- This is a 1-dimensional 

- All elements of an atomic vector must be the  same type, Hence it is a **homogeneous** type of object. Vectirs can hold numeric data, charactor data or logical data. 

### Creating vectors

Vectors can be created by using the function concatenation `c`


**Syntax**
```r
vector_name <- c(element1, element2, element3)

```

**Examples**

```{r, eval=TRUE}
first_vec <- c(10, 20, 50, 70)
second_vec <- c("Jan", "Feb", "March", "April")
third_vec <- c(TRUE, FALSE, TRUE, TRUE)
fourth_vec <- c(10L, 20L, 50L, 70L)
```

### Types and tests with vectors

1. `typepf()` returns types of their elements


```{r}
typeof(first_vec)
typeof(fourth_vec)
```


2. To check if it is a 

- vector: `is.vector()`

```{r}
is.vector(first_vec)
```

- charactor vector: `is.charactor()`

```{r}
is.character(first_vec)
```

- double: `is.double()`

```{r}
is.double(first_vec)
```

- integer: `is.integer()`

```{r}
is.integer(first_vec)
```

- logical: `is.logical()`

```{r}
is.logical(first_vec)
```

- atomic: `is.atomic()`

```{r}
is.atomic(first_vec)
```


3. `length()` returns number of elements in a vector

```{r}
length(first_vec)
length(fourth_vec)
```

### Coercion

Vectors must be homogeneous. When you attempt to combine different types they will be coerced to the most flexible type so that every element in the vector is of the same type.

Order from least to most flexible

`logical` --> `integer` --> `double` --> `charactor`

```{r}
a <- c(3.1, 2L, 3, 4, "GPA") 
typeof(a) 
```

```{r}
anew <- c(3.1, 2L, 3, 4)
typeof(anew) 
```


### Explicit coercion

Vectors can be explicitly coerced from one class to another using the functions `as.charactor`, `as.numeric`, `as.integer`, and `as.logical`.

```{r}
vec1 <- c(TRUE, FALSE, TRUE, TRUE)
typeof(vec1)

vec2 <- as.integer(vec1)
typeof(vec2)
vec2
```

**Question**

Why the below output produce NAs?

```{r}
x <- c("a", "b", "c")
as.numeric(x)
```

### Simplifying vector creation

1. colon `:` produce regular spaced ascending or descending sequences.

```{r}
a1 <- 10:16
a1
b1 <- -0.5:8.5
b1
```

2. sequence `seq()`. There are three arguments we need to provide, i) initial value, ii) final value, and iii) increment

syntax

```r
seq(initial_value, final_value, increment)
```

example

```{r}


```

3. repeats `rep()` 

```{r}
rep(9, 5)

rep(1:4, 2)

rep(1:4, each=2) # each element is repeated twice

rep(1:4, times=2) # whole sequence is repeated twice

rep(1:4, each=2, times=3)

rep(1:4, 1:4)

rep(1:4, c(4, 1, 4, 2))

```

### Logical operations

### Subsetting 

There are situations where we want to select only some of the elements of a vector. Following codes show various ways to select part of a vector object.

```{r}
data <- c(10, 20, 103, 124, 126)

data[1] # shows the first element 

data[-1] # shows all except the first item

data[1:3] # shows first three elements

data[c(1, 3, 4)]

data[data > 3]

data[data<20|data>120]

```

Example: How do you replace the 3rd element in the data vector by 203?

```{r}
data[3] <- 203
data
```

### Vector arithmetic

Vector operations are perfored element by element. 

```{r}
c(10, 100, 100) + 2 # two is added to every element in the vector
```

Vector operations between two vectors

```{r}
v1 <- c(1, 2, 3)
v2 <- c(10, 100, 1000)
v1 + v2
```

Add two vectors of unequal length

```{r}
longvec <- seq(10, 100, length=10)
shortvec <- c(1, 2, 3, 4, 5)

shortvec+longvec

```

### Missing values

Use `NA` to place a missing value in a vector.

```{r}
z <- c(10, 101, 2, 3, NA)
is.na(z)

```

## Matrix

Matrix is a 2-dimentional and a homogeneous data structure

**Syntax to create a matrix**

```r
matrix_name <- matrix(vector_of_elements, 
                      nrow=number_of_rows,
                      ncol=number_of_columns,
                      byrow=logical_value, # If byrow=TRUE, then the matrix is filled in by row.
                      dimnames=list(rnames, cnames)) # To assign row names and columns

```

**Example**

```{r}
values <- c(10, 20, 30, 40)
matrix1 <- matrix(values, nrow=2) # Matrix filled by columns (default option)
matrix1
matrix2 <- matrix(values, nrow=2, byrow=TRUE) # Matrix filled by rows
matrix2
```

**Naming matrix rows and columns**

```{r}
rnames <- c("R1", "R2")
cnames <- c("C1", "C2")
matrix_with_names <- matrix(values, nrow=2, dimnames=list(rnames, cnames))
matrix_with_names

```

### Matrix subscript

`matraix_name[i, ]` gives the ith row of a matrix

```{r}
matrix1[1, ]
```

`matraix_name[, j]` gives the jth column of a matrix

```{r}
matrix1[, 2]
```

`matraix_name[i, j]` gives the ith row and jth column element

```{r}
matrix1[1, 2]
```

```{r}
matrix1[1, c(1, 2)] 
```

### `cbind` and `rbind`

Matrices can be created by column-binding and row-binding with `cbind()` and `rbind()`

```{r}

x <- 1:3
y <- c(10, 100, 1000)

cbind(x, y) # binds matrices horizontally

rbind(x, y) #binds matrices vertically

```

### Matrix operations



## Array

- 3 dimentional data structure

- 


## List



## Data frame

- A data frame is more general than a matrix in that different columns can contain different modes of data.

- It’s similar to the datasets you’d typically see in SPSS and MINITAB.

- Data frames are the most common data structure you’ll deal with in R.

### Creating a dataframe

**Syntax**

```r
name_of_the_dataframe <- data.frame(
                          var1_name=vector of values of the first variable,
                          var2_names=vector of values of the second variable)

```

**Example**

```{r}
corona <- data.frame(ID=c("C001", "C002", "C003", "C004"),
                     Location=c("Beijing", "Wuhan", "Shanghai", "Beijing"),
                     Test_Results=c(FALSE, TRUE, FALSE, FALSE))
corona
```


To check if it is a datafrme

```{r}
is.data.frame(corona)

```

To convert a matrix to a dataframe

```{r}
mat <- matrix(10:81, ncol=4)
mat
mat_df <- as.data.frame(mat)
mat_df
```

### Subsetting data frames 

**Select rows**

```{r}
head(mat_df) # default it shows 5 rows

head(mat_df, 3) # To extract only the first three rows 
```

```{r}
tail(mat_df)
```

To select some specific rows

```{r}
index <- c(1, 3, 7, 8)
mat_df[index, ]

```

**Select columns**

1. Select column(s) by variable names

```{r}
mat_df$V1 # Method 1

mat_df[, "V1"] # Method 2
```

2. Select column(s) by index

```{r}
mat_df[, 2]
```

### Built in dataframes

**Note:**  All objects in R have a class.

