# Functions in R {#function}

A function is a block of organized and reusable code that is used to perform a specific task in a program. There are two types of functions in R:

  1. In-built functions
    
  2. User-defined functions
    
## In-built functions

These functions in R programming are provided by R environment for direct execution, to make our work easier Some examples for the frequently used in-built functions are as follows.

```{r comment=NA}
mean(c(10, 20, 21, 78, 105))

```


## User-defined functions

These functions in R programming language are dclared and defined by a user according to the requirements, to perform a specific task.

## Main components of a function

All R functions have three main components: (Check this with Hadley's book)

1. **function name**: name of the function that is stored as an R object

2. **arguments:** are used to rovide specific inputs to a function while a function is invoked. A function can have zero, single, multiple or default arguments.

3. **function body:** contains the block of code that performs the specific task assigned to a function. **return value** 

## Passing arguments to a function

## Some useful built-in functions in R


### R can be used as a simple calculator.

|  Operator | Description  |
|---|---|
| + | addition   |
| -  | substraction|
|*| multiplication|
|^|exponentiation (5^2 is 25)|
|%%| modulo-remainder of the division of the number to the left by the number on its right. (5%%3 is 2)|

### Some more maths functions

|  Operator | Description  |
|---|---|
|  abs(x) | absolute value of x   |
| log(x, base=y)  |  logarithm of x with base y; if base is not specified, returns the natural logarithm |
|exp(x)| exponential of x|
|sqrt(x)|square root of x|
|factorial(x)| factorial of x|

### Basic statistic functions

|  Operator | Description  |
|---|---|
|  mean(x) | mean of x   |
| median(x)  |  median of x |
|mode(x)| mode of x|
|var(x)|variance of x|
|scale(x)| z-score of x|
|quantile(x)| quantiles of x|
|summary(x)|summary of x: mean, minimum, maximum, etc.|

### Probability distribution functions

- **d** prefix for the **distribution** function

- **p** prefix for the **cummulative probability**

- **q** prefix for the **quantile**

- **r** prefix for the **random** number generator

#### Illustration with Standard normal distribution

The general formula for the probability density function of the normal distribution with mean $\mu$ and variance $\sigma$ is given by


\[
f_X(x) = \frac{1}{\sigma\sqrt{(2\pi)}} e^{-(x-\mu)^2/2\sigma^2}
\]

If we let the mean $\mu=0$ and the standard deviation $\sigma=1$, we get the probability density function for the standard normal distribution.

\[
f_X(x) = \frac{1}{\sqrt{(2\pi)}} e^{-(x)^2/2}
\]




```{r, comment=NA}
dnorm(0)
```

```{r, echo=FALSE, fig.cap="Standard normal probability density function: dnorm(0)", collapse=NA}
library(ggplot2)

x <- seq(-4, 4, length=200)
y <- 1/sqrt(2*pi)*exp(-x^2/2)
#plot(x, y, type="l", lwd=2, col="red")

dat <- data.frame(x=x, y=y)
ggplot(data = dat, mapping = aes(x = x, y = y)) +
    geom_line()+labs(y = "density", title = "dnorm(0)")+
    geom_segment(aes(x = 0, y = 0, xend = 0, yend = 0.398), size=1, color="red")+
    geom_segment(aes(x = 0, y = 0.398, xend = -4, yend = 0.398), arrow = arrow(length = unit(0.5, "cm")), size=1, color="red")
    

```


```{r, comment=NA}
pnorm(0)
```

```{r, echo=FALSE, fig.cap="Standard normal probability density function: dnorm(0)", collapse=NA}
library(ggplot2)

x <- seq(-4, 4, length=200)
y <- 1/sqrt(2*pi)*exp(-x^2/2)
#plot(x, y, type="l", lwd=2, col="red")

dat <- data.frame(x=x, y=y)
ggplot(data = dat, mapping = aes(x = x, y = y)) +
    geom_line()+labs(y = "density", title = "pnorm(0)")+
    geom_area(stat = "function", fun = dnorm, fill = "red", xlim = c(-4, 0))
    

```

```{r, comment=NA}
qnorm(0.5)
```

```{r, echo=FALSE, fig.cap="Standard normal probability density function: dnorm(0)", collapse=NA}
library(ggplot2)

x <- seq(-4, 4, length=200)
y <- 1/sqrt(2*pi)*exp(-x^2/2)
#plot(x, y, type="l", lwd=2, col="red")

dat <- data.frame(x=x, y=y)
ggplot(data = dat, mapping = aes(x = x, y = y)) +
    geom_line()+labs(y = "density", title = "qnorm(0)")+
    geom_segment(aes(x = 0, y = 0.398, xend = 0, yend = 0), size=1, color="red", arrow = arrow(length = unit(0.5, "cm")),)+
    geom_segment(aes(x = 0, y = 0.398, xend = -4, yend = 0.398), size=1, color="red")
    

```


```{r, comment=NA}
set.seed(262020)
random_numbers <- rnorm(10)
random_numbers
sort(random_numbers) ## sort the numbers then it is easy to map with the graph

```


```{r, echo=FALSE}
x <- seq(-4, 4, length=200)
y <- 1/sqrt(2*pi)*exp(-x^2/2)
dat <- data.frame(x=x, y=y)
highlight <- data.frame(y=rep(0, 10), x=random_numbers)
ggplot(data = dat, mapping = aes(x = x, y = y)) +
    geom_line() + labs(y = "density", title = "rnorm(10)") + geom_point(data=highlight, aes(x=x, y=y), colour="red", size=2)

```


### Reproducibility of scientific results

```{r, comment=NA}
rnorm(10) # first attempt

```


```{r, comment=NA}
rnorm(10) # second attempt

```

As you can see above you will get different results 

```{r, comment=NA}
set.seed(1)
rnorm(10) # First attempt with set.seed
```

```{r, comment=NA}
set.seed(1)
rnorm(10) # Second attempt with set.seed
```



